An Environment for Sustainable Research Software in Germany and Beyond: Current State, Open Challenges, and Call for Action

### 2020-04-03 manuscript.tex
Add final PDF

%%% Journal toggle; only specific options recognised.
%%% (Only "gigascience" and "general" are implemented now. Support for other journals is planned.)
\reviewfrom{23 January 2020}
\reviewto{PLACEHOLDER}
\acceptedon{PLACEHOLDER}

@@

% Custom packages
\usepackage{soulutf8} % TODO Remove
\usepackage{blindtext} % TODO Remove
\usepackage[nonumberlist]{glossaries} % TODO Check if we can put glossary items in the abbreviations section
%\usepackage[inline]{enumitem}
% Set graphics path
\graphicspath{{figures/}}
% EO Custom packages

@@

\title{A template for de-RSE position papers}

%%% Use the \authfn to add symbols for additional footnotes, if any. 1 is reserved for correspondence emails; then continuing with 2 etc for contributions.
\author[1,2,3,\authfn{1},\authfn{2}]{Stephan Druskat}
\author[4,5,6,\authfn{2}]{Frank L\"offler}
\author[7]{Kim Author}

\affil[1]{Department of English Studies, Friedrich Schiller University Jena, Germany}
\affil[2]{Institute of Simulation and Software Technology, German Aerospace Center (DLR), Germany}
\affil[3]{Department of Computer Science, Humboldt-Universit\"at zu Berlin, Germany}
\affil[4]{Friedrich Schiller University Jena, Germany}
\affil[5]{Michael Stifel Center Jena, Germany}
\affil[6]{Center for Computation and Technology, Louisiana State University, Baton Rouge, USA}
\affil[7]{University of Arcadia, Erewhon}

%%% Author Notes
\authnote{\authfn{1}vorstand@de-rse.org; \authfn{2}Contributed equally.}

@@

\newglossaryentry{GeopoliticalUnits}
{
    name = {geopolitical units},
    description = {Governed public units, ranging from cities and councils, over federal states and countries, up to political unions such as the EU. In the context of this paper, the discussion usually focuses on the larger units (countries and political unions)}
}
\newglossaryentry{RSE}
{
    name = {Research Software Engineers (RSEs)},
    description = {People creating and maintaining research software; this group ranges from research-focused software developers, to software engineers with a focus on research; other definitions include other roles, such as research software managers}
}
\newglossaryentry{DomainResearchers}
{
    name = {domain researchers},
    description = {The people doing the research to advance knowledge in a field}
}

@@

\begin{frontmatter}
\maketitle
\begin{abstract}
    Lorem ipsum dolor sit amet, consectetur adipiscing elit. In faucibus malesuada nisi a viverra. Pellentesque id nisi commodo, vulputate libero in, blandit nisl. Aenean sed enim erat. Duis at lorem non enim dignissim fringilla. In varius posuere sapien ut luctus. Pellentesque porttitor aliquam sapien, quis vestibulum dolor condimentum vitae. Aliquam congue sem at nibh vulputate blandit. Donec in tortor ac lectus consequat tempus quis sodales nunc. Vestibulum id nisi pellentesque, maximus neque quis, facilisis elit. Donec interdum lorem congue, rhoncus dui id, scelerisque purus. Cras sed lectus quis risus convallis hendrerit. Cras in ipsum pretium, mattis sapien eu, hendrerit augue. Duis metus neque, pulvinar et lectus in, molestie mollis sapien. Sed non dolor sed odio sagittis semper aliquam quis metus. Etiam risus augue, pharetra scelerisque egestas at, tempor id leo. 
\end{abstract}

\begin{keywords}
Semicolon-separated keywords; Lots of them
\end{keywords}
\end{frontmatter}

@@

\section{Background}
\label{sec:background}

Some background. Lorem ipsum dolor sit amet, consectetur adipiscing elit. In faucibus malesuada nisi a viverra. Pellentesque id nisi commodo, vulputate libero in, blandit nisl. Aenean sed enim erat. Duis at lorem non enim dignissim fringilla. In varius posuere sapien ut luctus. Pellentesque porttitor aliquam sapien, quis vestibulum dolor condimentum vitae. Aliquam congue sem at nibh vulputate blandit. Donec in tortor ac lectus consequat tempus quis sodales nunc. Vestibulum id nisi pellentesque, maximus neque quis, facilisis elit. Donec interdum lorem congue, rhoncus dui id, scelerisque purus. Cras sed lectus quis risus convallis hendrerit. Cras in ipsum pretium, mattis sapien eu, hendrerit augue. Duis metus neque, pulvinar et lectus in, molestie mollis sapien. Sed non dolor sed odio sagittis semper aliquam quis metus. Etiam risus augue, pharetra scelerisque egestas at, tempor id leo.

Vivamus aliquet quis lorem a cursus~\cite{druskatSoftwareDependenciesResearch2019}. Curabitur dapibus molestie pharetra. Aenean semper orci eu libero elementum, id ultricies est varius. Interdum et malesuada fames ac ante ipsum primis in faucibus. Nullam convallis, eros id congue lacinia, turpis eros consectetur felis, non sodales arcu nisl iaculis risus. In placerat, elit ac pellentesque posuere, dolor justo tincidunt felis, id condimentum velit est eu enim. Ut vitae tristique nibh. Praesent luctus eleifend volutpat.

Duis in tempor libero, at malesuada felis. Curabitur laoreet rutrum dui sed commodo. Curabitur vel eros massa. Sed ultricies lacinia nulla, ut imperdiet urna dignissim eget. Nam sit amet felis blandit, vulputate tortor in, semper dui. Pellentesque lorem sem, tincidunt eu sapien eget, finibus vulputate lacus. Cras ut condimentum est. Praesent vehicula aliquet tincidunt. Pellentesque vel mauris et nisl aliquet elementum. Ut ut tristique sem. Nullam pretium ex efficitur nulla ultricies malesuada. Praesent et maximus nisi. In venenatis risus ac orci molestie, quis tincidunt ligula eleifend.

Ut quis semper lacus, non hendrerit nisl. Quisque scelerisque tristique libero eu iaculis. Ut eleifend leo consectetur magna egestas gravida. Praesent pellentesque massa et consectetur auctor. Nullam tincidunt nisi id eros dignissim, vel ultrices erat molestie. Sed sit amet tristique ipsum. Fusce tempus, eros vitae lacinia interdum, nunc ligula scelerisque mi, ut euismod augue sem id sapien. Fusce imperdiet dui vitae lorem scelerisque, vitae consectetur dolor egestas.

Suspendisse nec fringilla massa. Nullam accumsan mi non erat elementum, sed imperdiet ligula posuere. In euismod, magna id consequat ornare, arcu dolor imperdiet arcu, a mollis tortor libero non lorem. Sed mi massa, eleifend et tristique id, hendrerit sed tellus. Nunc tincidunt, ex at pellentesque porttitor, lacus erat vulputate lectus, et tincidunt ligula odio quis leo. Vestibulum ante ipsum primis in faucibus orci luctus et ultrices posuere cubilia Curae; Etiam faucibus blandit nisi, et accumsan ipsum fermentum eu. Duis pulvinar tempus ligula, in condimentum erat commodo in. Sed ultrices vitae justo a sagittis. Phasellus condimentum odio ligula, et suscipit tellus ullamcorper eu. Vestibulum nec volutpat eros, et porta turpis. Nam tincidunt risus a nisi venenatis volutpat.

\subsection{A subsection}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. In faucibus malesuada nisi a viverra. Pellentesque id nisi commodo, vulputate libero in, blandit nisl. Aenean sed enim erat. Duis at lorem non enim dignissim fringilla. In varius posuere sapien ut luctus. Pellentesque porttitor aliquam sapien, quis vestibulum dolor condimentum vitae. Aliquam congue sem at nibh vulputate blandit. Donec in tortor ac lectus consequat tempus quis sodales nunc. Vestibulum id nisi pellentesque, maximus neque quis, facilisis elit. Donec interdum lorem congue, rhoncus dui id, scelerisque purus. Cras sed lectus quis risus convallis hendrerit. Cras in ipsum pretium, mattis sapien eu, hendrerit augue. Duis metus neque, pulvinar et lectus in, molestie mollis sapien. Sed non dolor sed odio sagittis semper aliquam quis metus. Etiam risus augue, pharetra scelerisque egestas at, tempor id leo.

\subsubsection[]{A subsubsection}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. In faucibus malesuada nisi a viverra. Pellentesque id nisi commodo, vulputate libero in, blandit nisl. Aenean sed enim erat. Duis at lorem non enim dignissim fringilla. In varius posuere sapien ut luctus. Pellentesque porttitor aliquam sapien, quis vestibulum dolor condimentum vitae. Aliquam congue sem at nibh vulputate blandit. Donec in tortor ac lectus consequat tempus quis sodales nunc. Vestibulum id nisi pellentesque, maximus neque quis, facilisis elit. Donec interdum lorem congue, rhoncus dui id, scelerisque purus. Cras sed lectus quis risus convallis hendrerit. Cras in ipsum pretium, mattis sapien eu, hendrerit augue. Duis metus neque, pulvinar et lectus in, molestie mollis sapien. Sed non dolor sed odio sagittis semper aliquam quis metus. Etiam risus augue, pharetra scelerisque egestas at, tempor id leo.

\section{Conclusions}
% -- Hartwig: Please check whether I include all aspects in the rewritten conclusions. I tried to make it more concise, and give clear advice to the readers (funding bodies)
Lorem ipsum. Just lorem ipsum.

>>>>

%%% Journal toggle; only specific options recognised.
%%% (Only "gigascience" and "general" are implemented now. Support for other journals is planned.)
\reviewfrom{23 January 2020}
\reviewto{09 February 2020}
\acceptedon{03 April 2020}

@@

% Custom packages
\usepackage[nonumberlist]{glossaries} % TODO Check if we can put glossary items in the abbreviations section
%\usepackage[inline]{enumitem}
% Set graphics path
\graphicspath{{figures/}}
% EO Custom packages

@@

\title{An Environment for Sustainable Research Software in Germany and Beyond: Current State, Open Challenges, and Call for Action}

%%% Use the \authfn to add symbols for additional footnotes, if any. 1 is reserved for correspondence emails; then continuing with 2 etc for contributions.
\author[1,2,\authfn{2}]{Hartwig Anzt} % 0000-0003-2177-952X
\author[1,\authfn{2}]{Felix Bach} % 0000-0002-5035-7978
\author[3,4,5,\authfn{2}]{Stephan Druskat} % 0000-0003-4925-7248
\author[6,7,8,\authfn{2}]{Frank L\"offler} % 0000-0001-6643-6323
\author[9,\authfn{1},\authfn{2}]{Axel Loewe} % 0000-0002-2487-4744
\author[10,11\authfn{2}]{Bernhard Y.\ Renard} % 0000-0003-4589-9809
\author[12,13,\authfn{1}\authfn{2}]{Gunnar Seemann} % 0000-0001-7111-7992
\author[14,\authfn{2}]{Alexander Struck} % 0000-0002-1173-9228
\author[15]{Elke Achhammer}
\author[16]{Piush Aggarwal}
\author[17]{Franziska Appel}
\author[18]{Michael Bader}
\author[19]{Lutz Brusch}
\author[20]{Christian Busse}
\author[21]{Gerasimos Chourdakis}
\author[22]{Peter Ebert}
\author[23]{Bernd Flemisch}
\author[24]{Sven Friedl}
\author[25]{Bernadette Fritzsch}
\author[26]{Maximilian D.\ Funk}
\author[3]{Volker Gast}
\author[27]{Florian Goth}
\author[28]{Jean-No\"{e}l Grad}
\author[29]{Sibylle Hermann}
\author[30]{Florian Hohmann}
\author[31]{Stephan Janosch}
\author[32]{Dominik Kutra}
\author[33]{Jan Linxweiler}
\author[10,34]{Thilo Muth}
\author[35]{Wolfgang Peters-Kottig}
\author[36]{Fabian Rack}
\author[37]{Fabian H.C.\ Raters}
\author[38]{Stephan Rave}
\author[39]{Guido Reina}
\author[40]{Malte Reißig}
\author[41,42]{Timo Ropinski}
\author[43]{Joerg Schaarschmidt}
\author[44,45,46]{Heidi Seibold}
\author[47]{Jan P.\ Thiele}
\author[48]{Benjamin Uekerman}
\author[49]{Stefan Unger}
\author[28]{Rudolf Weeber}
\author[50]{Piotr Wojciech Dabrowski}

\affil[1]{Steinbuch Centre for Computing, Karlsruhe Institute of Technology (KIT), Germany}
\affil[2]{Innovative Computing Lab, University of Tennessee, Knoxville, TN, USA}
\affil[3]{Department of English Studies, Friedrich Schiller University Jena, Germany}
\affil[4]{Institute of Software Technology, German Aerospace Center (DLR), Germany}
\affil[5]{Department of Computer Science, Humboldt-Universit\"at zu Berlin, Germany}
\affil[6]{Heinz Nixdorf Chair for Distributed Information Systems, Friedrich Schiller University Jena, Germany}
\affil[7]{Michael Stifel Center Jena, Germany}
\affil[8]{Center for Computation and Technology, Louisiana State University, Baton Rouge, USA}
\affil[9]{Institute of Biomedical Engineering, Karlsruhe Institute of Technology (KIT), Germany}
\affil[10]{Bioinformatics Unit (MF1), Robert Koch Institute, Berlin, Germany}
\affil[11]{Hasso Plattner Institute, Faculty of Digital Engineering, University of Potsdam, Germany}
\affil[12]{Institute for Experimental Cardiovascular Medicine, University Heart Centre Freiburg Bad Krozingen, Germany}
\affil[13]{Faculty of Medicine, University of Freiburg, Freiburg, Germany}
\affil[14]{Matters of Activity. Image Space Material. Cluster of Excellence at Humboldt-Universität zu Berlin, Germany}
\affil[15]{Technische Universität München, Germany}
\affil[16]{Language Technology Lab, Universität Duisburg-Essen, Germany}
\affil[17]{Leibniz Institute of Agricultural Development in Transition Economies (IAMO), Halle (Saale), Germany}
\affil[18]{Department of Informatics, Technical University of Munich
, Germany}
\affil[19]{Center for Information Services and High Performance Computing (ZIH), Technische Universität Dresden, Germany}
\affil[20]{Deutsches Krebsforschungszentrum, Heidelberg, Germany}
\affil[21]{Chair of Scientific Computing in Computer Science, Technical University Munich, Germany}
\affil[22]{Center for Bioinformatics Saar, Saarland Informatics Campus, Germany}
\affil[23]{Institute for Modelling Hydraulic and Environmental Systems, University of Stuttgart, Germany}
\affil[24]{Berlin Institute of Health, Germany}
\affil[25]{Scientific Computing, Alfred Wegener Institute, Helmholtz Center for Polar and Marine Research Bremerhaven, Germany}
\affil[26]{Max-Planck-Gesellschaft e.V.}
\affil[27]{Institut für Theoretische Physik und Astrophysik, Universität Würzburg, Germany}
\affil[28]{Institut für Computerphysik, University of Stuttgart, Germany}
\affil[29]{University Library, University of Stuttgart, Germany}
\affil[30]{Zentrum für Medien-, Kommunikations- und Informationsforschung (ZeMKI), Universität Bremen, Germany}
\affil[31]{Max Planck Institue of Molecular Cell Biology and Genetics, Dresden, Germany}
\affil[32]{European Molecular Biology Laboratory, Heidelberg, Germany}
\affil[33]{Center for Mechanics, Uncertainty and Simulation in Engineering, Technische Universität Braunschweig, Germany}
\affil[34]{eScience Division, Federal Institute for Materials Research and Testing, Berlin, Germany}
\affil[35]{Konrad-Zuse-Zentrum für Informationstechnik Berlin (ZIB), Germany}
\affil[36]{FIZ Karlsruhe - Leibniz Institute for Information Infrastructure, Germany}
\affil[37]{Department of Economics, University of Goettingen, Germany}
\affil[38]{Applied Mathematics, University of Münster, Germany}
\affil[39]{Visualization Research Center, University of Stuttgart, Germany}
\affil[40]{Institute for Advanced Sustainability Studies e.V.}
\affil[41]{Institute of Media Informatics, Ulm University, Germany}
\affil[42]{Department of Science and Technology, Linköping University, Sweden}
\affil[43]{Institute of Nanotechnology, Karlsruhe Institute of Technology (KIT),  Germany}
\affil[44]{LMU Munich, Germany}
\affil[45]{Bielefeld University, Germany}
\affil[46]{Helmholtz Zentrum Munich, Germany}
\affil[47]{Institute of Applied Mathematics, Leibniz University Hannover, Germany}
\affil[48]{Energy Technology, Eindhoven University of Technology, The Netherlands}
\affil[49]{Julius Kühn-Institut (JKI), Federal Research Centre for Cultivated Plants, Quedlinburg, Germany}
\affil[50]{School of Computing, Communication and Business, Hochschule für Technik und Wirtschaft Berlin, Germany}

%%% Author Notes
\authnote{\authfn{1}axel.loewe@kit.edu, gunnar.seemann@universitaets-herzzentrum.de; \authfn{2}Contributed equally.}

@@

\newglossaryentry{GeopoliticalUnits}
{
    name = {geopolitical units},
    description = {Governed public units, ranging from cities and councils, over federal states and countries, up to political unions such as the EU. In the context of this paper, the discussion usually focuses on the larger units (countries and political unions)}
}
\newglossaryentry{RSE}
{
    name = {research software engineers (RSEs)},
    description = {People creating and maintaining research software; this group ranges from research-focused software developers, to software engineers with a focus on research; other definitions include other roles, such as research software managers}
}
\newglossaryentry{DomainResearchers}
{
    name = {domain researchers},
    description = {The people doing the research to advance knowledge in a field}
}

@@

\begin{frontmatter}
\maketitle
\begin{abstract}
Research software has become a central asset in academic research.
It optimizes existing and enables new research methods, implements and embeds research knowledge,
and constitutes an essential research product in itself.
Research software must be sustainable in order to understand, replicate, reproduce, and build upon existing research or conduct new research effectively. In other words, software must be available, discoverable, usable, and adaptable to new needs, both now and in the future.
Research software therefore requires an environment that supports sustainability.

Hence, a change is needed in the way research software development and maintenance are currently
motivated, incentivized, funded, structurally and infrastructurally supported, and legally treated.
Failing to do so will threaten the quality and validity of research.
In this paper, we identify challenges for research software sustainability
in Germany and
beyond, in terms of motivation, selection, research software engineering personnel, funding, infrastructure, and legal aspects.
Besides researchers, we specifically address political and academic decision-makers to increase awareness of the importance and needs of sustainable research software practices. In particular, we recommend strategies and measures to create an environment for sustainable research software,
with the ultimate goal
to ensure that software-driven research is valid, reproducible and sustainable, and that software is recognized as a first class citizen in research.
This paper is the outcome of two workshops run in Germany in 2019, at deRSE19 - the first International Conference of Research Software Engineers in Germany - and a dedicated DFG-supported follow-up workshop in Berlin.
\end{abstract}

\begin{keywords}
Sustainable Software Development; Academic Software; Software Infrastructure; Software Training; Software Licensing; Research Software
\end{keywords}
\end{frontmatter}

@@

\section{Background}
\label{sec:background}
\noindent\vsepfbox{
  \parbox[]{\dimexpr\columnwidth-3\fboxsep-2\fboxrule\relax}{
    Meet Kim, who is currently a post-grad PhD student in researchonomy at the University of Arcadia (UofA). We will follow Kim’s fictional career in order to understand different aspects of research software sustainability. Note that in Kim's world, many of the changes this paper calls for have already been implemented.
    \\
    {\footnotesize (In our example, Kim is a female person. Of course, research software engineers (RSEs) can be any gender.)}
  }
}

\noindent Computational analysis of large data sets, computer-based simulations, and software technology in general play a central role for virtually all scientific breakthroughs of at least the 21st century. The first image of a black hole may be the most prominent recent example where astrophysical experiments and the collection and processing of data had to be complemented with sophisticated algorithms and software to enable research excellence~\cite{%TheEventHorizonTelescopeCollaboration2019,TheEventHorizonTelescopeCollaboration2019a,TheEventHorizonTelescopeCollaboration2019b,
TheEventHorizonTelescopeCollaboration2019c,Nowogrodzki2019%,TheEventHorizonTelescopeCollaboration2019d,TheEventHorizonTelescopeCollaboration2019e
}. Similarly, it is research software that allows us to get a glimpse of the consequences our actions today have on the climate of tomorrow. However, an implication of computer-based research is that findings and data can only be reproduced, understood, and validated if the software that was used in the research process is sustained and their functionality maintained.

At the same time, sustaining research software, and in particular open research software, comes with a number of challenges. Commercial research software often has revenue flows that can facilitate sustainable software development, maintenance, and documentation as well as the operation of adequate infrastructure. However, a large share of researchers base their research on software that was developed in-house or as a community effort. Many of these software stacks can not be sustained -- often because research software was not a first class deliverable in a research project and hence remained in a prototype state, or because of missing incentives and resources to maintain the software after project funding ended.
Another fundamental difference to industrial software development is that most developers of academic research software (often doctoral students or postdoctoral researchers) never receive training in sustainable software development~\cite{SSIanalysis}. In particular, as they see themselves usually as the primary user of a software product, there are virtually no incentives to invest in sustainability measures such as code documentation or portability.
In combination with the predominance of temporary positions in research, this results in a highly inefficient system where millions of lines of code are generated every year that will not be re-used after the termination of the developer's position.
Part of the problem is the reluctance to accept research software engineering as an academic profession that results in a lack of incentives to produce high-quality software: producing high software quality needs sufficient resources, and although the San Francisco Declaration on Research Assessment (DORA~\cite{DORA}) demands a change in the academic credit system, many institutions base promotion and appointments on traditional metrics like the Hirsch index~\cite{Hirsch2005}.
It is obvious that an extraordinary amount of idealism is required to write sustainable code including documentation and installation routines, as well as running infrastructure and giving support to others when resources can be used more profitably in writing scientific publications based on fragile prototype software~\cite{bangerth2014quo,Prins2015}.

Thus, one main factor for the poor sustainability of research software is the lack of long-term funding for research software engineers (RSEs)~\cite{RSEdef} who take care of the appropriate architecture, organization, implementation, documentation, and community interaction for the software, paired with the implementation of measures towards making the software sustainable during and beyond the development process~\cite{SSIRSE}.

In this paper, we describe the state of the practice and current challenges for research software sustainability, and suggest measures towards improvements that can solve these challenges.
The paper is the result of a community effort, with work undertaken during two workshops and subsequent collaborative work across the larger RSE community in Germany.
It has been initiated during a half-day workshop at the first International Conference for Research Software Engineers in Germany (deRSE19) in Potsdam, Germany on June 5th, 2019~\cite{deRSEconf}, and continued during a dedicated two-day workshop in Berlin, Germany on November 7th and 8th, 2019, which was funded by the German Research Foundation (Deutsche Forschungsgemeinschaft, DFG).
Subsequently, the draft produced during the latter event was opened up for collaborative discussion by the German RSE community through \textit{de-RSE e.V. - Society for Research Software}.
%Public reviews were conducted between PLACEHOLDER and PLACEHOLDER.
%This version of the paper has also been accepted as an official position paper of de-RSE.

We mainly focus on the situation of research software and RSEs in Germany, where funding bodies increasingly acknowledge the importance and value of sustainable research software and related infrastructures.
The DFG, the largest funding body for fundamental research in Germany, for example, opened a call for sustainable research software development~\cite{DFGcall1} at the end of 2016 and a second call for quality management in research software~\cite{DFGcall2} in June 2019. The first call was oversubscribed by a factor of 10-15, a strong indicator of unmet demand. As another example, the 2019 ``Guidelines for Safeguarding Good Research Practice'' codex of the DFG~\cite{DFGcode} now explicitly lists software side-by-side with other research results and data. The FAIR principles for research data~\cite{Wilkinson2016} provide guidelines for data archiving, but enabling full reproducibility and traceability of research software requires additional steps~\cite{hasselbring2019fair}. In consequence, there are ongoing discussions on whether software should be considered as a specific kind of research data or as a separate entity~\cite{lamprechttowards}.

These positive developments notwithstanding, guidelines and policies for sustainable research software
development in Germany are unfortunately still lacking, and long-term funding
strategies are missing. This all leads to unmet requirements and
unsolved challenges that we want to highlight in this paper by
elaborating on
\begin{inlinelist}
    \item why research software engineering needs to be considered an integral part of academic research;
    \item how to decide which software to sustain; 
    \item who sustains research software;
    \item how software can be funded sustainably;
    \item what infrastructure is needed for sustainable software development; and
    \item legal aspects of research software development in academia.
\end{inlinelist}
While we specifically focus on the research software landscape in
Germany, we are convinced that many of the analyses, findings, and
recommendations may carry beyond. We want to address RSEs who are
experiencing similar challenges and newcomers to the field of research
software development, but first and foremost political and academic decision makers to
raise awareness of the importance of and requirements for sustainable software
development. As a community we work hard on overcoming the challenges of
software development in an academic setting, but we need support -- and reliable funding options and institutional recognition in particular -- for
the sake of better research.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Why Sustainable Research Software in the First Place?}
\label{sec:motivation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Stephan Druskat (Moderation, stephan.druskat@dlr.de), Florian Goth (fgoth@physik.uni-wuerzburg.de), Jean-Noël Grad (jgrad@icp.uni-stuttgart.de), Dominik Kutra (dominik.kutra@embl.de), Thilo Muth (mutht@rki.de), Fabian H. C. Raters (raters@uni-goettingen.de), Guido Reina (guido.reina@visus.uni-stuttgart.de)
\noindent\vsepfbox{
  \parbox[]{\dimexpr\columnwidth-3\fboxsep-2\fboxrule\relax}{
    After graduation, Kim joins a fixed-term researchonomical research project. For her PhD thesis, she wants to crunch some data. Her colleague recommends learning some Boa, which is an all-purpose programming language often used in researchonomy. Luckily, the UofA runs regular Software Plumbery courses for researchers, including a Boa course. Kim takes the course and gains a solid understanding of the basics of the Hash shell, version control with Tig, and the basics of Boa. She starts writing scripts, which help her a lot with the data processing. Unfortunately, Kim’s scripts are quite slow and actually break after she installs a newer version of Boa. She visits the weekly Code Café organized by her university’s central RSE team. The RSEs not only help her update her scripts but also suggest some changes which speed up the computation by a factor of 25.
    \\
    During the next meeting with her PhD supervisor, Kim presents her collection of scripts. The supervisor encourages Kim to create a Boa library from them, as they will be very useful to other researchonomists. Thankfully, Kim’s project PI had applied for 3 RSE person months in their grant, so the project enlists an RSE from the central team. Over the next three months, Kim and the RSE work together to build the library, document it, test it, license it under the permissive Comanche license, update the TigLab repository to let others contribute, introduce automated builds for every code change via a continuous integration platform, and make the library citable. Finally, they release the first major version of the library, named \emph{hal9k} and publish it through the university library’s software portal, where they get a DOI (Digital Object Identifier) for the version as well as a concept DOI for any future versions of the library. Working with the RSE, Kim has gained a good understanding of some methods in software engineering, and she’s thrilled because this also means she’ll be able to get a job with a local tech company once her fixed-term contract has run out.
    \\
    Kim passes her PhD - of which \emph{hal9k} is an important part - with flying colors, and soon citations to her library start appearing in the researchonomic literature. To Kim’s surprise, she also reads a blog post about a citizen science maker project which has used \emph{hal9k} to process researchonomic data measured in a neighborhood of her hometown. She is invited to give a talk at the local office of Siren, a global tech company, which look to adopt \emph{hal9k}, and pay Kim a generous speaker honorarium. So generous in fact, that Kim can pay a student assistant for a full year from the money.
  }
}
Our credibility as researchers in society hinges on the notion of proper
research conduct, also known as ``good research practice''. The
digitalization of research has introduced complex digital research
outputs, such as software and data sets. Although first
recommendations~\cite{katerbow2018} 
and policies~\cite{https://doi.org/10.2312/os.helmholtz.003} exist, they are far from being widely adopted. It is still somewhat unclear how to translate good research practice into good research
software practice, for example in terms
of validity and reproducibility, but also pertaining to the responsible use of resources. The damage
that failing to do so is causing both to the progress of the research community
and to the credibility of academic research in society is becoming increasingly clear
with the growth of the replication crisis - while the lack of universally
agreed-upon and supported good research software practice is not the main
reason for that crisis, it clearly is a contributing factor.

While it is obvious that software qualifies as a potentially re-usable
digital artifact, the additional benefit of not just reproducing a given
scenario, but transferring software use to new problems, domains,
and/or applications, justifies developing research software with a long-term perspective as \emph{sustainable research software}.

In order to support research, a sustainable software must be correct~\cite{hattonChimeraSoftwareQuality2007,changRetraction2006,matthewsFiveRetractedStructure2007}, validatable, understandable, documented, publicly
released, adequately published (i.e. in persistently identifiable form
as software source code~\cite{smithSoftwareCitationPrinciples2016}, and
potentially in an additional paper which describes the software concept, design decisions, and development rationale),
actively maintained, and \mbox{(re-)usable}~\cite{meraliComputationalScienceError2010,barnesPublishYourComputer2010,tseComputerCodeMore2010}.
We also argue that truly sustainable research software must ideally be published under a Free/Libre Open Source Software (FLOSS) license, and follow an open development model, to (1) enable the validation of research results that have been produced using the software, (2) enable the reproducibility of software-based research, (3) enable improvement and (re-) use of the software to support more and better research, and reduce resources to be spent on software development, (4) reduce legal issues (see section below), (5) meet ethical obligations from public funding, and (6) open research software to the general public, i.e., the stakeholder group with arguably the greatest interest in furthering research knowledge and improving research for the benefit of all.

To make software-based research (and with that almost any research)
reproducible, the used software must continue to exist. Furthermore, it must
continue to be usable, understandable, and return
consistent results (or potential changes to results and bug fixes must be clearly documented) in the
evolving software and hardware environment. Moreover, the software
should support reuse scenarios to avoid duplication of efforts and
unneeded drain of resources. Therefore, if research software is publicly
funded, it should be freely available under a FLOSS license.

Currently, creating and using sustainable research software is not sufficiently incentivized. To evaluate in which area this shortcoming should be addressed, we have identified the following challenges: \label{sec:motivation:challenges}

\begin{itemize}
\item
  \textbf{Lack of benefit for the individual}:
  Currently, the primary motivation for
    sustainable research software is the common benefit, rather than the
    individual benefit. It is  clearly beneficial for the
    research community as a whole to direct resources towards sustainable
    research software, as it enables better and more
    research by freeing funds for domain research rather than
    (repetitive) software development. But the developers are often even at a disadvantage (e.g., they publish fewer papers~\cite{bangerth2014quo,Prins2015}), which
    in turn prevents sustainable research software.
\item
  \textbf{Lack of suitable incentive systems}: Contributions to research that are not
  traditional text-based products (i.e., papers or monographs) are still not sufficiently rewarded, or not rewarded at all, due to the missing implementation of mandatory software citation ~\cite{smithSoftwareCitationPrinciples2016,Hafer2009,Howison2016,Li2017,Li2019,Park2019,Pan2019,Doerr2019,druskatSoftwareDependenciesResearch2019,katzSoftwareCitationImplementation2019}, among other reasons. Interestingly, one third of research software repositories have a lifespan (defined as the time from the first time
any code was uploaded to the last contribution) of less than one day (median: 15 days~\cite{hasselbring2019fair}), indicating that many codes are only made available publicly for the publication in a journal (as increasingly encouraged or required by journals~\cite{Resnik2019139} and associated with higher impact~\cite{codeImpact}) but are not maintained thereafter. 
\item
  \textbf{Lack of awareness}: Research software sustainability (see~\cite{Venters2014,Goble2014,Druskat2016,KatzFundamentalsSoftwareSustainability2018}) and its
  importance is lacking visibility as well as acceptance, and research software
  engineering in its implementation as sustainable software development
  and software maintenance is not sufficiently supported, both in
  Germany and beyond~\cite{SSIRSE,NWOprofile,Casties2019}.
\item
  \textbf{Lack of expertise}: Knowledge about how to create, maintain, and
  support sustainable research software is emerging~\cite{Wilson2014,Stodden2014,Wilson2017} but
  has not yet permeated related activities within organizations - specifically
  teaching, mentoring, and consultancy. This lack of expertise can also lead to divergence between
  software design and community uptake, e.g., if the software fails to meet the needs of the
  target group, or is insufficiently usable. RSEs
  combine sustainable software engineering expertise with experience in one or
  more research domains.
\item
  \textbf{Heterogeneous research community}:
  There are significant differences with respect to how software is developed, published, used, and valued in the different academic disciplines. Additionally, there is even heterogeneity within a community in terms of application and approach.
  This also makes it hard to train researchers for sustainable software development,
  as beyond basic training in computational research such as provided by The Carpentries, advanced courses for research software engineering are not widely available (with the notable exception of the CodeRefinery project~\cite{CodeRefinery}).
  Targeted curricula must be developed and updated regularly, and specialized instructors need to be trained.
\item
  \textbf{Lack of impact measures}: It is unclear how to measure the impact
  of research software with respect to its quality, reusability, and benefit for the research community. This exceeds the
  implementation of research software citation (which is work in progress~\cite{smithSoftwareCitationPrinciples2016,liSoftwareCitationReuse2016,katzSoftwareCitationImplementation2019,druskatSoftwareDependenciesResearch2019}), and pertains to sustainability and policy studies.
\item
  \textbf{Infrastructure issues}: Due to a lack of knowledge about
  how sustainability features impact the application of research software, there is not yet enough evidence
  for whether centralized or decentralized facilities should be favored to
  further research software sustainability~\cite{Kuchinke2016,SuLMaSS,Hexatomic}.
  This in turn leads to a lack of infrastructure as a whole.
\item
  \textbf{Legal issues}: Many obstacles for research software pertain to
  legal issues, such as applicable licensing and compatibility of licenses~\cite{morinQuickGuideSoftware2012}, and decisions about license types.
\item
  \textbf{Funding issues}: Despite some individual initiatives~\cite{katzLookingSoftwareSustainability2015,DFGcall1,DFGcall2,ChanZuckerberg}, funding for the creation, maintenance, and
  support of sustainable research software is still scarce.
\item
  \textbf{Slow adoption of research software engineering as a profession}: Career options for
  research software work are not fully determined, although career paths are emerging in some regions.
  Initially, the RSE initiative in the UK has made progress in this area, and RSE groups have been installed in many institutions. In Germany, the
  US, and the Netherlands, this is still work in progress~\cite{SSIRSE,de-rse,us-rse,nl-rse}. It is also not yet
  determined how to match research software engineering roles in public
  institutions with industry roles (see~\cite{rodriguez-sanchezAcademiaFailureRetain2017}).
\end{itemize}

In summary, the necessary but resource-intensive practice of creating, maintaining, supporting, and funding sustainable
research software is not yet sufficiently incentivized and enabled by
research institutions and funding agencies, nor does it align well with
the publish-or-perish culture that is still prominent in most fields.

Therefore, it is necessary to comprehensively motivate sustainable research software
practice.
In the following, we identify stakeholders of research software (see~\cite{Katz2019,Druskat2018,ye2019open}), and explicate their particular motivations for sustainable research software. Subsequently, we
specify challenges towards satisfying the demands of the individual stakeholders.

\subsection{Stakeholder Motivations for Research Software Sustainability}

While a wide range of stakeholders share interest in sustainable software, we argue that their individual motivation can differ quite significantly:

The \textit{\gls{GeneralPublic}} benefits from research which supports the
common good, in other terms: creates a better world, faster. Taxpayers
have an interest in economical use of their tax money, to which
duplicated or flawed efforts to create research software -- in contrast to software reuse -- is contrary. A subset of this group may be interested in sustainable, i.e., re-usable and understandable, software as part of citizen
science.
